---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pedigreedata

<!-- badges: start -->
[![R-CMD-check](https://github.com/R-Computing-Lab/pedigreedata/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/R-Computing-Lab/pedigreedata/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/R-Computing-Lab/pedigreedata/graph/badge.svg)](https://app.codecov.io/gh/R-Computing-Lab/pedigreedata)
<!-- badges: end -->

`pedigreedata` is an umbrella data package that provides curated pedigree and phenotype datasets for use in quantitative genetics and behavior genetics research. Each dataset has been validated for pedigree integrity and is ready for use with packages such as [BGmisc](https://github.com/R-Computing-Lab/BGmisc) and [ggpedigree](https://github.com/R-Computing-Lab/ggpedigree).

## Datasets

| Dataset | Description | Individuals | Source |
|---------|-------------|-------------|--------|
| `guinea_pigs` | Four-generation controlled breeding study of *Cavia porcellus* with repeated body weight measurements (days 0–90) | `r nrow(pedigreedata::guinea_pigs)` | [Figshare](https://doi.org/10.6084/m9.figshare.31513204.v2) |
| `red_squirrels` | 35+ years of pedigree and fitness data for wild North American red squirrels (*Tamiasciurus hudsonicus*) from the Kluane Red Squirrel Project | `r nrow(pedigreedata::red_squirrels)` | [Dryad](https://doi.org/10.5061/dryad.n5q05) |
| `war_of_the_roses` | Pedigree of key figures in the 15th-century English Wars of the Roses, rooted at Edward III | `r nrow(pedigreedata::war_of_the_roses)` | Public genealogical records |

## Installation

You can install the development version of pedigreedata from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("R-Computing-Lab/pedigreedata")
```

## Quick start

```{r example}
library(pedigreedata)

# Guinea pig growth data
head(guinea_pigs[, c("ID", "dadID", "momID", "sexo", "famID", "p0", "p90")])

# Red squirrel fitness data
head(red_squirrels[, c("personID", "momID", "dadID", "sex", "byear", "lrs")])

# Wars of the Roses pedigree
head(war_of_the_roses[, c("id", "name", "sex", "momID", "dadID", "famID")])
```

## Guinea pigs

`guinea_pigs` combines a four-generation pedigree with body weights measured at
six ages (days 0, 15, 30, 45, 60, and 90). Key variables include litter size,
season of birth (*invierno* / *verano*), coat color, and housing pen.

```{r gp-growth, message=FALSE, warning=FALSE, fig.height=4}
library(ggplot2)
library(dplyr)
library(tidyr)

growth_long <- guinea_pigs |>
  filter(!is.na(p0)) |>
  select(ID, sexo, p0, p15, p30, p45, p60, p90) |>
  pivot_longer(starts_with("p"), names_to = "day", values_to = "weight") |>
  mutate(day = as.numeric(sub("p", "", day)))

set.seed(42)
ids <- sample(unique(growth_long$ID), 800)

ggplot(
  growth_long |> filter(ID %in% ids),
  aes(x = day, y = weight, group = ID)
) +
  geom_line(alpha = 0.08) +
  geom_smooth(aes(group = sexo, color = sexo), method = "loess", se = TRUE) +
  scale_x_continuous(breaks = c(0, 15, 30, 45, 60, 90)) +
  scale_color_brewer(
    palette = "Set1",
    labels = c("H" = "Female", "M" = "Male"),
    na.value = "grey50"
  ) +
  labs(
    title = "Guinea pig growth trajectories (days 0–90)",
    x = "Day", y = "Weight (g)", color = "Sex"
  ) +
  theme_minimal()
```

## Red squirrels

`red_squirrels` spans over 35 years of individual-level data from a wild boreal
population. Fitness traits include lifetime reproductive success (LRS) and mean
annual reproductive success (ARS).

```{r squirrel-lrs, fig.height=3.5}
red_squirrels |>
  filter(!is.na(lrs), !is.na(sex)) |>
  ggplot(aes(x = lrs, fill = sex)) +
  geom_histogram(binwidth = 1, alpha = 0.75, position = "identity") +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Lifetime reproductive success by sex",
    x = "LRS", y = "Count", fill = "Sex"
  ) +
  theme_minimal()
```

## War of the Roses

`war_of_the_roses` encodes the competing dynastic claims of Lancaster and York.
Each individual has a Wikipedia URL, making it straightforward to trace lineages.

```{r wor-table}
# Key claimants and their relationship to Edward III (id = 1)
war_of_the_roses |>
  filter(name %in% c(
    "Edward III", "Henry IV", "Henry V", "Henry VI",
    "Richard Duke of York", "Edward IV", "Richard III",
    "Henry VII"
  )) |>
  select(id, name, sex, momID, dadID) |>
  arrange(id)
```

## Vignettes

Each dataset has a dedicated vignette:

- `vignette("guinea-pig", package = "pedigreedata")` — growth trajectories, litter structure, seasonal effects
- `vignette("red-squirrels", package = "pedigreedata")` — lifespan, LRS/ARS distributions, birth cohort trends
- `vignette("wars-of-the-roses", package = "pedigreedata")` — dynastic lineages and succession rules

## License

GPL (>= 3)
