---
title: "Guinea Pigs: A four-generation breeding study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guinea Pigs: A four-generation breeding study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

# Overview

The `guinea_pigs` dataset records pedigree and body weight measurements from
a four-generation controlled breeding program of domestic guinea pigs
(*Cavia porcellus*). Guinea pigs — known as *cuy* in Andean countries — are
an important livestock species in South America. This dataset captures
`r nrow(pedigreedata::guinea_pigs)` individuals, each with up to six body
weight measurements (days 0, 15, 30, 45, 60, and 90), along with pedigree
links, litter information, coat color, season of birth, and housing pen.

```{r load}
library(pedigreedata)
library(ggplot2)
library(dplyr)
library(tidyr)
library(BGmisc)
data(guinea_pigs)
```

---

# The pedigree
## Validate the pedigree

Before exploring the data, we validate the pedigree structure using
[BGmisc](https://github.com/R-Computing-Lab/BGmisc). `checkSex()` standardizes
sex coding and flags inconsistencies; `checkParentIDs()` identifies individuals
listed as parents who have no row of their own (adding phantom placeholders
where needed); `checkPedigreeNetwork()` confirms the pedigree is acyclic —
no individual can be their own ancestor.

```{r validate-sex}
gp <- checkSex(guinea_pigs,
  code_male   = "M",
  code_female = "H",
  verbose     = TRUE,
  repair      = TRUE
)
```

```{r validate-parents}
gp <- checkParentIDs(gp,
  addphantoms       = FALSE,
  repair            = TRUE,
  parentswithoutrow = FALSE,
  repairsex         = FALSE
)
```

```{r validate-network}
net <- checkPedigreeNetwork(gp,
  personID = "ID",
  momID    = "momID",
  dadID    = "dadID",
  verbose  = TRUE
)
cat("Pedigree is acyclic:", net$is_acyclic, "\n")
```

With the pedigree confirmed acyclic, we assign family group IDs using
`ped2fam()`. Family groups are the connected components of the pedigree
graph — sets of individuals linked by at least one parent-offspring
relationship.

```{r fam-ids}
gp <- ped2fam(gp, personID = "ID", famID = "famID")
cat("Number of family groups:", n_distinct(gp$famID), "\n")
```



## Who is in the data?

```{r pedigree-overview}
cat("Individuals:        ", nrow(guinea_pigs), "\n")
cat(
  "Founders:           ",
  sum(is.na(guinea_pigs$dadID) & is.na(guinea_pigs$momID)), "\n"
)
cat(
  "With both parents:  ",
  sum(!is.na(guinea_pigs$dadID) & !is.na(guinea_pigs$momID)), "\n"
)
cat("Generations:        ", paste(sort(unique(guinea_pigs$gener)), collapse = ", "), "\n")
```

Most individuals have both parents recorded — this is a controlled breeding
study, not a wild population, so pedigree completeness is high.

## Pedigree depth across generations

The breeding program ran for four generations. Each generation was larger
than the last as the population expanded.

```{r generation-sex, fig.height = 3.5}
guinea_pigs |>
  filter(!is.na(gener), !is.na(sex)) |>
  count(gener, sex) |>
  ggplot(aes(x = factor(gener), y = n, fill = sex)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(
    palette = "Set1",
    labels = c("H" = "Female", "M" = "Male")
  ) +
  labs(
    title = "Number of individuals per generation",
    x = "Generation", y = "Count", fill = "Sex"
  ) +
  theme_minimal()
```

## Litter structure

Guinea pigs are polytocous — they give birth to litters. Litter size at
birth (`tc_nac`) and at weaning (`tc_destete`) reflect both reproductive
output and early-life survival.

```{r litter-size, fig.height = 3.5}
guinea_pigs |>
  filter(!is.na(tc_nac)) |>
  count(tc_nac) |>
  ggplot(aes(x = factor(tc_nac), y = n)) +
  geom_col(fill = "#377EB8", alpha = 0.85) +
  labs(
    title = "Litter size at birth",
    x = "Litter size (tc_nac)", y = "Count"
  ) +
  theme_minimal()
```

```{r litter-survival}
guinea_pigs |>
  filter(!is.na(tc_nac), !is.na(tc_destete)) |>
  mutate(survived = tc_destete / tc_nac) |>
  summarise(
    mean_born   = round(mean(tc_nac), 2),
    mean_weaned = round(mean(tc_destete), 2),
    mean_surv   = round(mean(survived), 3)
  )
```

On average, most offspring that are born survive to weaning — consistent
with the controlled conditions of a breeding program.

## Family group sizes

```{r fam-sizes, fig.height = 3.5}
guinea_pigs |>
  count(famID) |>
  ggplot(aes(x = n)) +
  geom_histogram(binwidth = 5, fill = "#377EB8", alpha = 0.85) +
  labs(
    title = "Distribution of family group sizes",
    x = "Individuals per family", y = "Number of families"
  ) +
  theme_minimal()
```

## A single family pedigree

```{r single-family, eval = requireNamespace("ggpedigree", quietly = TRUE), fig.width = 8, fig.height = 5}
library(ggpedigree)

fam_summary <- guinea_pigs |>
  filter(!is.na(momID) | !is.na(dadID)) |>
  count(famID) |>
  filter(n >= 15, n <= 40) |>
  arrange(desc(n))

focal_fam <- guinea_pigs |> filter(famID == fam_summary$famID[1])

ggPedigree(focal_fam,
  personID = "ID",
  momID = "momID",
  dadID = "dadID",
  config = list(
    add_phantoms = TRUE,
    code_male    = "M",
    code_female  = "H"
  )
)
```

---

# Growth trajectories

## Overall growth

Body weight was recorded at six time points across the first 90 days of
life. Weights increase rapidly in the first 30 days, then plateau.

```{r growth-plot}
growth_long <- guinea_pigs |>
  filter(!is.na(p0)) |>
  select(ID, sex, p0, p15, p30, p45, p60, p90) |>
  pivot_longer(
    cols      = starts_with("p"),
    names_to  = "day",
    values_to = "weight"
  ) |>
  mutate(day = as.numeric(stringr::str_remove(day, "p")))

set.seed(42)
sample_ids <- sample(unique(growth_long$ID), 1000)

ggplot(
  growth_long |> filter(ID %in% sample_ids),
  aes(x = day, y = weight, group = ID)
) +
  geom_line(alpha = 0.1) +
  geom_jitter(alpha = 0.05, width = 1, height = 0) +
  geom_smooth(
    method = "loess", se = TRUE,
    aes(group = sex, color = sex)
  ) +
  scale_x_continuous(breaks = c(0, 15, 30, 45, 60, 90)) +
  scale_color_brewer(
    palette  = "Set1",
    labels   = c("H" = "Female", "M" = "Male"),
    na.value = "grey50"
  ) +
  labs(
    title = "Growth trajectories over the first 90 days",
    x     = "Day",
    y     = "Weight (g)",
    color = "Sex"
  ) +
  theme_minimal()
```

Male and female growth curves diverge after day 30, with males reaching
higher weights by day 90. This sex difference is a consistent feature of
guinea pig growth and should be included as a covariate in any analysis.

## Weight at day 90 by sex

```{r p90-summary}
guinea_pigs |>
  filter(!is.na(p90), !is.na(sex)) |>
  group_by(Sex = sex) |>
  summarise(
    n    = n(),
    mean = round(mean(p90), 1),
    sd   = round(sd(p90), 1),
    min  = min(p90),
    max  = max(p90)
  )
```

---

# Seasonal variation

The dataset records whether each animal was born in winter (*invierno*) or
summer (*verano*). Season of birth may influence early resource availability,
dam condition, and growth.

## Birth weight by season

```{r season-birthweight, fig.height = 3.5}
guinea_pigs |>
  filter(!is.na(p0), !is.na(clima)) |>
  ggplot(aes(x = p0, fill = clima)) +
  geom_histogram(binwidth = 15, alpha = 0.7, position = "identity") +
  scale_fill_manual(
    values = c("invierno" = "#377EB8", "verano" = "#E41A1C"),
    labels = c("invierno" = "Winter", "verano" = "Summer")
  ) +
  labs(
    title = "Birth weight by season",
    x = "Birth weight (g)", y = "Count", fill = "Season"
  ) +
  theme_minimal()
```

```{r season-summary}
guinea_pigs |>
  filter(!is.na(p0), !is.na(clima)) |>
  group_by(Season = clima) |>
  summarise(
    n            = n(),
    mean_birthwt = round(mean(p0, na.rm = TRUE), 1),
    mean_p90     = round(mean(p90, na.rm = TRUE), 1)
  )
```

## Season effect on growth trajectories

Does the season of birth shape how animals grow across the full 90 days,
or is the effect limited to birth weight?

```{r season-growth}
growth_season <- guinea_pigs |>
  filter(!is.na(p0), !is.na(clima)) |>
  select(ID, clima, p0, p15, p30, p45, p60, p90) |>
  pivot_longer(
    cols      = starts_with("p"),
    names_to  = "day",
    values_to = "weight"
  ) |>
  mutate(day = as.numeric(stringr::str_remove(day, "p")))

set.seed(42)
sample_ids2 <- sample(unique(growth_season$ID), 1000)

ggplot(
  growth_season |> filter(ID %in% sample_ids2),
  aes(x = day, y = weight, group = ID)
) +
  geom_line(alpha = 0.08) +
  geom_smooth(
    method = "loess", se = TRUE,
    aes(group = clima, color = clima)
  ) +
  scale_x_continuous(breaks = c(0, 15, 30, 45, 60, 90)) +
  scale_color_manual(
    values = c("invierno" = "#377EB8", "verano" = "#E41A1C"),
    labels = c("invierno" = "Winter", "verano" = "Summer")
  ) +
  labs(
    title = "Growth trajectories by season of birth",
    x     = "Day",
    y     = "Weight (g)",
    color = "Season"
  ) +
  theme_minimal()
```

---

# Other characteristics

## Coat color

```{r coat-color, fig.height = 3}
guinea_pigs |>
  filter(!is.na(color), color != "", color != "na") |>
  count(color, sort = TRUE) |>
  ggplot(aes(x = reorder(color, n), y = n)) +
  geom_col(fill = "#4DAF4A", alpha = 0.85) +
  coord_flip() +
  labs(
    title = "Coat color distribution",
    x = NULL, y = "Count"
  ) +
  theme_minimal()
```

## Missing data

Not all individuals were weighed at every time point — attrition increases
at later ages.

```{r missingness}
weight_cols <- c("p0", "p15", "p30", "p45", "p60", "p90")
sapply(guinea_pigs[weight_cols], function(x) {
  paste0(sum(!is.na(x)), " (", round(mean(!is.na(x)) * 100, 1), "%)")
}) |>
  setNames(paste0("Day ", c(0, 15, 30, 45, 60, 90)))
```

---

# Source

Figshare: <https://doi.org/10.6084/m9.figshare.31513204.v2>
