---
title: "Guinea Pig Pedigree and Growth Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guinea Pig Pedigree and Growth Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load}
library(pedigreedata)
```

## Overview

This vignette describes the guinea pig (*Cavia porcellus*) datasets included in
`pedigreedata`. The data come from a four-generation breeding study and are
available from Figshare
([doi:10.6084/m9.figshare.31513204.v2](https://doi.org/10.6084/m9.figshare.31513204.v2)).

Four related datasets are provided:

| Dataset | Description | Rows |
|---------|-------------|------|
| `ped` | Pedigree only (ID, sire, dam) | `r nrow(ped)` |
| `pheno` | Phenotype and weight measurements | `r nrow(pheno)` |
| `ped_pheno` | Full join of pedigree and phenotype | `r nrow(ped_pheno)` |
| `ped_growth` | Long-format growth trajectories | `r nrow(ped_growth)` |

---

## Pedigree structure (`ped`)

```{r ped-head}
head(ped)
```

```{r ped-summary}
# Founders: individuals with no recorded parents
founders <- ped[is.na(ped$dadID) & is.na(ped$momID), ]
cat("Total individuals:", nrow(ped), "\n")
cat("Founders (no parents recorded):", nrow(founders), "\n")
cat("Non-founders:", nrow(ped) - nrow(founders), "\n")
```

---

## Phenotype data (`pheno`)

### Structure

```{r pheno-head}
head(pheno)
```

```{r pheno-str}
str(pheno)
```

### Sample characteristics

```{r pheno-sex}
table(Sex = pheno$sexo, useNA = "ifany")
```

```{r pheno-season}
table(Season = pheno$clima, useNA = "ifany")
```

```{r pheno-color}
table(Color = pheno$color, useNA = "ifany")
```

```{r pheno-generation}
table(Generation = pheno$gener, useNA = "ifany")
```

### Weight measurements

```{r pheno-weights}
weight_cols <- c("p0", "p15", "p30", "p45", "p60", "p90")
weight_summary <- sapply(pheno[weight_cols], function(x) {
  c(
    n       = sum(!is.na(x)),
    missing = sum(is.na(x)),
    mean    = round(mean(x, na.rm = TRUE), 1),
    sd      = round(sd(x, na.rm = TRUE), 1),
    min     = min(x, na.rm = TRUE),
    max     = max(x, na.rm = TRUE)
  )
})
t(weight_summary)
```

---

## Growth trajectories (`ped_growth`)

`ped_growth` is a long-format version suitable for plotting or mixed models,
with one row per individual per measurement day.

```{r growth-head}
head(ped_growth)
```

```{r growth-plot, fig.width = 6, fig.height = 4}
# Random sample of 500 individuals to keep plot readable
set.seed(42)
sample_ids <- sample(unique(ped_growth$ID), 500)
plot_data <- ped_growth[ped_growth$ID %in% sample_ids, ]

# Remove rows with missing sex or weight
plot_data <- plot_data[!is.na(plot_data$sexo) & !is.na(plot_data$weight), ]

# Base R plot
palette(c("#E41A1C", "#377EB8"))
plot(
  weight ~ day,
  data = plot_data,
  col = adjustcolor(as.factor(plot_data$sexo), alpha.f = 0.07),
  pch = 16, cex = 0.6,
  xlab = "Day", ylab = "Weight (g)",
  main = "Growth trajectories (sample of 500 individuals)",
  axes = FALSE
)
axis(1, at = c(0, 15, 30, 45, 60, 90))
axis(2)
box()

# Loess smoother per sex
for (s in c("H", "M")) {
  d <- plot_data[plot_data$sexo == s, ]
  lo <- loess(weight ~ day, data = d, span = 0.5)
  day_seq <- seq(0, 90, by = 1)
  lines(day_seq, predict(lo, newdata = data.frame(day = day_seq)),
    col = as.integer(as.factor(s)), lwd = 2
  )
}
legend("topleft",
  legend = c("Female (H)", "Male (M)"),
  col = 1:2, lwd = 2, bty = "n"
)
```

---

## Joining pedigree and phenotype (`ped_pheno`)

`ped_pheno` is a full join of `ped` and `pheno`, retaining all individuals
regardless of whether they appear in both sources.

```{r ped-pheno-overlap}
in_both <- sum(!is.na(ped_pheno$p0))
ped_only <- sum(is.na(ped_pheno$p0) & is.na(ped_pheno$sexo))
cat("Individuals with phenotype data:  ", in_both, "\n")
cat("Individuals in pedigree only:      ", nrow(ped_pheno) - in_both, "\n")
```

---

## Citation

If you use these data, please cite the original source:

> Figshare dataset. <https://doi.org/10.6084/m9.figshare.31513204.v2>
