---
title: "Red Squirrels: 35 years of a wild population"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Red Squirrels: 35 years of a wild population}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

# Overview

The `red_squirrels` dataset comes from the **Kluane Red Squirrel Project**,
a long-term field study of North American red squirrels (*Tamiasciurus
hudsonicus*) in the boreal forests of Yukon, Canada, running continuously
since 1987. Every squirrel in the population is individually marked, and
its parentage, birth year, death year, and reproductive output are
recorded across its lifetime.

The result is one of the most complete wild-animal pedigrees ever assembled:
`r nrow(pedigreedata::red_squirrels)` individuals spanning over three
decades, with fitness measures for most of them.

```{r load}
library(pedigreedata)
library(ggplot2)
library(dplyr)

data(red_squirrels)
head(red_squirrels)
```

---

# The study population over time

## When were squirrels born?

The project began marking squirrels in 1987. Birth year coverage shows
how the monitored population grew — and which years had the most recruits.

```{r birth-years, fig.height = 3.5}
red_squirrels |>
  filter(!is.na(byear)) |>
  count(byear) |>
  ggplot(aes(x = byear, y = n)) +
  geom_col(fill = "#E41A1C", alpha = 0.8) +
  labs(
    title = "Number of squirrels born per year",
    x = "Birth year", y = "Count"
  ) +
  theme_minimal()
```

The spikes in some years reflect the boreal forest's **mast cycle** —
periodic high cone production that triggers a population boom in red
squirrels (a phenomenon called the "Chitty effect" in this system).

## Lifespan: birth to death

For individuals with both birth and death years recorded, we can look
at the distribution of lifespans.

```{r lifespan, fig.height = 3.5}
red_squirrels |>
  filter(!is.na(byear), !is.na(dyear)) |>
  mutate(lifespan = dyear - byear) |>
  filter(lifespan >= 0) |>
  ggplot(aes(x = lifespan, fill = sex)) +
  geom_histogram(binwidth = 1, alpha = 0.7, position = "identity") +
  scale_fill_brewer(
    palette = "Set1",
    labels = c("F" = "Female", "M" = "Male")
  ) +
  labs(
    title = "Lifespan distribution by sex",
    x = "Lifespan (years)", y = "Count", fill = "Sex"
  ) +
  theme_minimal()
```

```{r lifespan-summary}
red_squirrels |>
  filter(!is.na(byear), !is.na(dyear), !is.na(sex)) |>
  mutate(lifespan = dyear - byear) |>
  filter(lifespan >= 0) |>
  group_by(Sex = sex) |>
  summarise(
    n          = n(),
    mean_years = round(mean(lifespan), 2),
    max_years  = max(lifespan)
  )
```

---

# Pedigree structure

## Founders and family depth

```{r pedigree-overview}
cat("Total individuals:  ", nrow(red_squirrels), "\n")
cat(
  "Founders:           ",
  sum(is.na(red_squirrels$momID) & is.na(red_squirrels$dadID)), "\n"
)
cat(
  "With both parents:  ",
  sum(!is.na(red_squirrels$momID) & !is.na(red_squirrels$dadID)), "\n"
)
cat("Number of families: ", n_distinct(red_squirrels$famID), "\n")
```

## Family sizes

The dataset includes `r n_distinct(red_squirrels$famID)` distinct family
groups. Most are small (wild territorial animals rarely aggregate), but
a subset of well-connected lineages spans multiple generations.

```{r family-sizes, fig.height = 3.5}
red_squirrels |>
  count(famID) |>
  ggplot(aes(x = n)) +
  geom_histogram(binwidth = 2, fill = "#377EB8", alpha = 0.8) +
  labs(
    title = "Distribution of family group sizes",
    x = "Individuals per family", y = "Number of families"
  ) +
  theme_minimal()
```

## Validate the pedigree

Before plotting, we validate the pedigree structure. `checkSex()` standardizes
sex coding; `checkParentIDs()` adds phantom placeholders for parents with no
row of their own; `checkPedigreeNetwork()` confirms there are no impossible
ancestry loops.

```{r validate}
sq <- checkSex(red_squirrels,
  code_male   = "M",
  code_female = "F",
  verbose     = TRUE,
  repair      = TRUE
)

sq <- checkParentIDs(sq,
  addphantoms       = TRUE,
  repair            = TRUE,
  parentswithoutrow = FALSE,
  repairsex         = FALSE
)

net <- checkPedigreeNetwork(sq,
  personID = "ID",
  momID    = "momID",
  dadID    = "dadID",
  verbose  = TRUE
)
cat("Pedigree is acyclic:", net$is_acyclic, "\n")
```

## A single family pedigree

```{r single-family, eval = requireNamespace("ggpedigree", quietly = TRUE), fig.width = 8, fig.height = 5}
library(ggpedigree)

# Pick a mid-sized family with known parentage for a clean plot
fam_summary <- sq |>
  filter(!is.na(momID) | !is.na(dadID)) |>
  count(famID) |>
  filter(n >= 15, n <= 30) |>
  arrange(desc(n))

focal_fam <- sq |> filter(famID == fam_summary$famID[1])

ggPedigree(focal_fam,
  personID = "ID",
  momID = "momID",
  dadID = "dadID",
  config = list(
    add_phantoms = TRUE,
    code_male    = "M",
    code_female  = "F"
  )
)
```

---

# Fitness: how many offspring do squirrels produce?

## Lifetime reproductive success (LRS)

LRS is the total number of offspring an individual produced that survived
to independence across its entire life — the closest field equivalent of
Darwinian fitness.

```{r lrs-dist, fig.height = 3.5}
red_squirrels |>
  filter(!is.na(lrs), !is.na(sex)) |>
  ggplot(aes(x = lrs, fill = sex)) +
  geom_histogram(binwidth = 1, alpha = 0.7, position = "identity") +
  scale_fill_brewer(
    palette = "Set1",
    labels = c("F" = "Female", "M" = "Male")
  ) +
  labs(
    title = "Lifetime reproductive success (LRS) by sex",
    x = "LRS", y = "Count", fill = "Sex"
  ) +
  theme_minimal()
```

```{r lrs-summary}
red_squirrels |>
  filter(!is.na(lrs), !is.na(sex)) |>
  group_by(Sex = sex) |>
  summarise(
    n         = n(),
    mean_lrs  = round(mean(lrs), 2),
    median    = median(lrs),
    max_lrs   = max(lrs),
    prop_zero = round(mean(lrs == 0), 2)
  )
```

The right skew is a universal feature of fitness distributions: most
individuals produce few or no recruits, while a small number are
extraordinarily productive. Notice that a large proportion have LRS = 0.

## Annual reproductive success (ARS)

ARS captures productivity in a single year, averaged across all years
the individual was observed. It is less sensitive to lifespan differences
than LRS.

```{r ars-dist, fig.height = 3.5}
red_squirrels |>
  filter(!is.na(ars_mean), !is.na(sex)) |>
  ggplot(aes(x = ars_mean, fill = sex)) +
  geom_histogram(binwidth = 0.25, alpha = 0.7, position = "identity") +
  scale_fill_brewer(
    palette = "Set1",
    labels = c("F" = "Female", "M" = "Male")
  ) +
  labs(
    title = "Mean annual reproductive success (ARS) by sex",
    x = "Mean ARS", y = "Count", fill = "Sex"
  ) +
  theme_minimal()
```

## Does reproductive success change over the study period?

Population-level fitness can vary with environmental conditions. Here
we look at mean LRS by birth cohort.

```{r lrs-over-time, fig.height = 3.5}
red_squirrels |>
  filter(!is.na(byear), !is.na(lrs)) |>
  group_by(byear) |>
  summarise(mean_lrs = mean(lrs), n = n()) |>
  filter(n >= 10) |>
  ggplot(aes(x = byear, y = mean_lrs)) +
  geom_point(aes(size = n), alpha = 0.6, color = "#E41A1C") +
  geom_smooth(method = "loess", se = TRUE, color = "black") +
  scale_size_continuous(range = c(1, 5)) +
  labs(
    title = "Mean LRS by birth cohort (cohorts with n ≥ 10)",
    x = "Birth year", y = "Mean LRS", size = "Cohort size"
  ) +
  theme_minimal()
```

Variation in mean LRS across birth cohorts reflects the boom-bust dynamics
of this population, driven by the mast cycle and other environmental factors.

---

# Missing data

Not all individuals have complete records — founders by definition lack
parent IDs, and not all individuals have fitness data.

```{r missing}
data.frame(
  Variable = c("momID", "dadID", "byear", "dyear", "lrs", "ars_mean"),
  Missing = sapply(
    red_squirrels[c("momID", "dadID", "byear", "dyear", "lrs", "ars_mean")],
    function(x) {
      paste0(
        sum(is.na(x)), " (",
        round(mean(is.na(x)) * 100, 1), "%)"
      )
    }
  )
)
```

---

# Source

McFarlane, S.E., Boutin, S., Humphries, M.M., et al. (2015). Very low
levels of direct additive genetic variance in fitness and fitness components
in a red squirrel population. Dryad.
<https://doi.org/10.5061/dryad.n5q05>
