---
title: "Kluane Red Squirrel Project Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kluane Red Squirrel Project Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load}
library(pedigreedata)
```

## Overview

The `redsquirrels` and `redsquirrels_full` datasets contain pedigree and
fitness data from the **Kluane Red Squirrel Project**, a long-term field study
of North American red squirrels (*Tamiasciurus hudsonicus*) in Yukon, Canada,
running continuously since 1987.

Two versions of the dataset are provided:

| Dataset | Description | Rows |
|---------|-------------|------|
| `redsquirrels` | Families with at least one recorded father; core fitness columns | `r nrow(redsquirrels)` |
| `redsquirrels_full` | All individuals; full ARS summary statistics | `r nrow(redsquirrels_full)` |

**ARS** = Annual Reproductive Success (offspring recruited per year).
**LRS** = Lifetime Reproductive Success (total offspring recruited).

---

## Core dataset (`redsquirrels`)

```{r head}
head(redsquirrels)
```

```{r str}
str(redsquirrels)
```

### Sex distribution

```{r sex}
table(Sex = redsquirrels$sex, useNA = "ifany")
```

### Founders and pedigree depth

```{r founders}
founders <- redsquirrels[is.na(redsquirrels$momID) & is.na(redsquirrels$dadID), ]
cat("Total individuals: ", nrow(redsquirrels), "\n")
cat("Founders:          ", nrow(founders), "\n")
cat("Non-founders:      ", nrow(redsquirrels) - nrow(founders), "\n")
```

### Birth and death years

```{r years}
cat("Birth years: ", range(redsquirrels$byear, na.rm = TRUE), "\n")
cat("Death years: ", range(redsquirrels$dyear, na.rm = TRUE), "\n")
cat("Missing byear:", sum(is.na(redsquirrels$byear)), "\n")
cat("Missing dyear:", sum(is.na(redsquirrels$dyear)), "\n")
```

### Fitness measures

```{r fitness}
cat("--- Lifetime Reproductive Success (LRS) ---\n")
summary(redsquirrels$lrs)
cat("\n--- Mean Annual Reproductive Success (ARS) ---\n")
summary(redsquirrels$ars_mean)
```

```{r fitness-plot, fig.width = 6, fig.height = 4}
# LRS distribution by sex
lrs_f <- redsquirrels$lrs[redsquirrels$sex == "F" & !is.na(redsquirrels$lrs)]
lrs_m <- redsquirrels$lrs[redsquirrels$sex == "M" & !is.na(redsquirrels$lrs)]

breaks <- seq(0, max(redsquirrels$lrs, na.rm = TRUE) + 2, by = 2)
hist(lrs_f, breaks = breaks, col = adjustcolor("#E41A1C", 0.5),
     xlab = "Lifetime Reproductive Success", main = "LRS by sex",
     ylim = c(0, max(hist(lrs_f, breaks = breaks, plot = FALSE)$counts,
                     hist(lrs_m, breaks = breaks, plot = FALSE)$counts) * 1.1))
hist(lrs_m, breaks = breaks, col = adjustcolor("#377EB8", 0.5), add = TRUE)
legend("topright", legend = c("Female", "Male"),
       fill = adjustcolor(c("#E41A1C", "#377EB8"), 0.5), bty = "n")
```

---

## Full dataset (`redsquirrels_full`)

`redsquirrels_full` retains all individuals (including those in matrilineal-only
families) and adds year-by-year ARS summary statistics.

```{r full-head}
head(redsquirrels_full)
```

```{r full-extra}
cat("Additional columns in redsquirrels_full:\n")
setdiff(names(redsquirrels_full), names(redsquirrels))
```

```{r full-ars}
cat("ARS observation counts (ars_n):\n")
summary(redsquirrels_full$ars_n)
cat("\nObservation span (years):\n")
span <- redsquirrels_full$year_last - redsquirrels_full$year_first
summary(span[!is.na(span)])
```

---

## Citation

If you use these data, please cite the original Dryad deposit:

> McFarlane, S.E., et al. (2015). Kluane Red Squirrel Project pedigree and
> fitness data. Dryad. <https://doi.org/10.5061/dryad.n5q05>
