---
title: "Wars of the Roses Pedigree"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Wars of the Roses Pedigree}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r load}
library(pedigreedata)
```

## Overview

The `warsofroses` dataset is a pedigree of the English royal families involved
in the Wars of the Roses (1455–1487). The pedigree is rooted at
**Edward III** (reigned 1327–1377), whose descendants formed the rival houses
of Lancaster and York. It was processed using
[BGmisc](https://github.com/R-Computing-Lab/BGmisc) to assign family IDs and
check pedigree integrity.

```{r dim}
cat("Individuals:", nrow(warsofroses), "\n")
cat("Columns:    ", ncol(warsofroses), "\n")
```

## Structure

```{r head}
head(warsofroses[, c("id", "name", "sex", "momID", "dadID", "famID")])
```

```{r str}
str(warsofroses)
```

## Sex and founder status

```{r sex}
table(Sex = warsofroses$sex, useNA = "ifany")
```

```{r founders}
founders <- warsofroses[is.na(warsofroses$momID) & is.na(warsofroses$dadID), ]
cat("Founders (no parents recorded):", nrow(founders), "\n")
founders[, c("id", "name", "sex")]
```

## Twins

```{r twins}
twins <- warsofroses[!is.na(warsofroses$twinID), ]
if (nrow(twins) > 0) {
  twins[, c("id", "name", "sex", "twinID", "zygosity")]
} else {
  cat("No twins recorded in this dataset.\n")
}
```

## Family structure

```{r families}
fam_sizes <- table(warsofroses$famID)
cat("Number of family groups:", length(fam_sizes), "\n")
cat("Family sizes:\n")
print(fam_sizes)
```

## Wikipedia links

Each individual has a Wikipedia URL where available, making this dataset
useful for teaching about pedigree structure and historical genealogy.

```{r urls}
# Individuals with Wikipedia entries
has_url <- warsofroses[!is.na(warsofroses$url) & warsofroses$url != "", ]
cat(nrow(has_url), "of", nrow(warsofroses), "individuals have Wikipedia URLs\n")
head(has_url[, c("name", "url")], 10)
```

## Citation

This dataset was compiled from public genealogical records. Wikipedia URLs
are provided for reference. If using in published work, please cite the
`pedigreedata` package.
